---
title: 'Assignment: Population dynamics'
output:
  word_document: default
  html_document:
    code_folding: show
date: "20 February 2019"
---

Remember the frog matrix? Your first task is to investigate how changing the population vector affects the short term dynamics in the frog population:


```{r, eval = TRUE}
library(popbio) #load popbio package

#create an empty matrix and specify the dimensions (3 * 3) 
#The arguments nrow and ncol defines the number of rows and columns.
A <-matrix(nrow=3, ncol=3) 

#Add fecundity measures:
A[1,] <- c(0, 52, 279.5) #fill first row
#add survival probabilites:
A[2,] <- c(0.019, 0.25, 0) #fill second row
A[3,] <- c(0, 0.08, 0.43)  #fill third row

A # inspect your matrix
```

In the first case, we start with 100 individuals distributed over the different life-stages according to the stable stage distribution: 

```{r, eval = TRUE}
n0_1 <- stable.stage(A)*100 
n0_1
```


In the second case, we start with 100 individuals, all adults. 
```{r, eval = TRUE}
n0_2 <- c(0,0,100) 
n0_2
```


In the third case, start with 100 individuals, all pre-juveniles. 
```{r, eval = TRUE}
n0_3 <- c(100,0,0) 
n0_3
```

Track each scenario for 14 years (iterations=15). 
 
```{r, eval = TRUE}
N1 <- pop.projection(A, n0_1, 14+1) 
N1$pop.sizes 

N2 <- pop.projection(A, n0_2, 14+1) 
N2$pop.sizes 

N3 <- pop.projection(A, n0_3, 14+1) 
N3$pop.sizes
```

a) Why do you think that the population sizes differ so much?  

*The answer should include that individuals from the different stages dont contribute equally to future populations. In this case, adults contribute the most to future population sizes (adults have the highest reproductive value) and pre-juveniles contribute the least (have the lowest reproductive value).*  


b) If you would attempt a reintroduction program, would you release juveniles or adults?  

*Adults*


Your second task is to calculate the growth rate and sensitivities for a population of orcas. Orcas have expanded northwards in response to climate change. This is expected to result in trophic cascades in their new foraging areas because they feed on other marine predator.

Load the whale matrix from popbio:
```{r, eval = TRUE}
library(popbio) #load popbio package
data(whale)
whale
```

Calculate the long term deterministic lambda for the population:  

```{r}
lambda(whale)
```

c) What does this say about future population sizes?  

*Lambda is larger than = 1 which means that the population is increasing. Future population sizes are likely to increase*

d) Do you trust that using this measure would result in accurate estimates of future population sizes of this natural orca population? Why/why not?  

*It could perhaps be trusted in the short term because it is based on actual reproduction and survival probabilities in a real population.*
*Main reason not to trust it is that it doesn't take sochastic factors into account. Stochastic lambda could be lower than the deterministic lambda (which is already close to 1 [the population is decreasing when lambda < 1]). That is, it is likely that the population size could in fact decrease in the future.* 

Calculate sensitivities and elasticities for the transitions in the whale matrix. Plot the resulting sensitivity and elasticity matrices.  

```{r, eval = TRUE}
image2(sensitivity(whale))
```
```{r, eval = TRUE}
image2(elasticity(whale))
```


e) What do these values say about the importance of the different transitions in the matrix for the population growth rate? What would be the implications for management?

*Answer should include something in terms of:*

*The sensitivities say how sensitive the population growth rate is to a tiny change in each vital rate/matrix element.*

*(Elasticities are proportional sensitivities that allow us to compare the importance of changes in vital rates for the population growth rate among vital rates/matrix elements)*

*Survival of mature individuals has a large effect on population growth rate and so does survival of juveniles. I would focus my efforts on management that lead to an increased survival in these classes*

*(We get sensitivities for several impossible transitions. In theory this would indicate that if it were possible to increase transitions from e.g. post-repoductive to mature this would have a large effect on population growth rate.)* 
